The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2020-02-09 pkg/partykit/inst/guideliketest/draft/draft.Rnw
first revisions based on reviewer's comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2976 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 43a79bfd0fe422b02dc3879c75c6800ba912bd11

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref,dsfont}
\usepackage[all]{hypcap}
\usepackage{enumitem}
\usepackage{color,soul}

@@

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e., the task of selecting the best split variable in a given (sub)sample. In particular, the approach of the 
GUIDE algorithm is compared to the one of the CTree algorithm and the MOB algorithm
by investigating the building blocks of their testing strategies in which they differ:
(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Dichotomization of these residuals or scores.
(3)~Categorization of possible split variables.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

@@

%\newpage
\subsection{Pruning}
\label{sec:pruning}

@@

Note that there are further differences in the testing strategies between the three algorithms, e.g.,
using conditional vs.\ unconditional approximations of the null distributions. However, this difference
has relatively little influence compared to the other building blocks considered in detail here.
Moreover, both similarities and relative differences between these approaches have been previously
discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.

@@

% To illustrate the effects of the building blocks in the unifying framework presented in 
% Section~\ref{sec:unify} different combinations of them are investigated for the first
% simulation scenario (stump) in Sections~\ref{sec:scores} and~\ref{sec:3way}. 
% Then the testing strategies are evaluated on the more
% complex tree structure of the second scenario in Section~\ref{sec:tree}. 
% The corresponding results are shown in the following plots.

\newpage
\subsection[``Stump'' scenario: Residuals vs. full model scores]{``Stump'' scenario: Residuals vs.\ full model scores}
\label{sec:scores}

@@

However, in the scenario where only $\beta_1$ (but not the intercept $\beta_0$) is affected by the split,
the residual-based GUIDE approach has no power at all even for a true split at the median (top middle panel).
It is easily possible, though, to substantially mitigate this problem by using scores
(sensitive to changes in all parameters) rather than residuals only (sensitive to changes
in the conditional mean). The remaining difference between MOB/CTree and GUIDE+scores
is due to dichotomizing the scores at zero and due to categorizing the split variables
which are investigated in more detail in the following section.

@@

Linear models as presented in this study have been chosen as they are highly relevant in practice,
allow for simple illustrations, and theoretical insights are available for the testing strategies.
However, the results can be easily extended to a wide variety of other models where the introduced building
blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
changes in the parameters do not lead to shifts in the residuals (but in other components of the
scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
of a linear model only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor which in our case is $(1,0)^{\top}$.
Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
is of type $(0, \delta)^\top$ and thus orthogonal to the mean regressor. Due to this residual-based
tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
interactions, see \citealp{Doove+Dusseldorp+VanDeun:2014}) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.
%% FIXME: add reference to Seibold+Hothorn+Zeileis:2019
Testing strategies being capable of detecting all kinds of changes including those orthogonal to the 
mean are also employed in an extension of (generalised) linear model trees introduced by 
\cite{Seibold+Hothorn+Zeileis:2019}
which allows for selected model parameters to be kept fixed at a globally estimated value while only 
changes in the remaining parameters are considered for splitting.

@@

\section*{Acknowledgments}
Torsten Hothorn received funding from the Swiss National Science
Foundation, grant number 200021\_184603.

@@

For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks are listed in Table~\ref{tab:combinations_appendix}.
%Each suffix indicates a change of the original form of the considered testing strategy.
%In particular, it relates to the transformations of the data before applying a statistical
%test as represented by the different building blocks. In that way,
%``+res'' refers to using only residuals (instead of all model scores),
%``+scores'' to using all model scores (instead of only residuals),
%``+cat'' to applying a categorization of split variables,
%``+bin'' to applying a dichotomization of scores/residuals.

@@

>>>>

%% packages
\usepackage{amstext,amsfonts,amsmath,amssymb,bm,thumbpdf,lmodern,hyperref,dsfont}
\usepackage[all]{hypcap}
\usepackage{enumitem}
\usepackage{color,xcolor,soul}

@@

\fixme{``Why CTree, GUIDE, and MOB?''}

\hlc{Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e., the task of selecting the best split variable in a given (sub)sample. 
For this purpose, three leading unbiased recursive partitioning algorithms are considered, 
namely GUIDE~(Google Scholar: 484), CTree~(2444) and MOB~(421). Moreover, through this 
selection the approach of QUEST (1306) is included as well due to GUIDE having been introduced 
as a descendant/extension of QUEST (and others). %\citep{} User Manual for GUIDE ver. 31.0.
In particular, GUIDE, CTree, and MOB are compared by investigating the building blocks of 
their testing strategies in which they differ:}
(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Dichotomization of these residuals or scores.
(3)~Categorization of possible split variables.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

@@

\subsection{Selecting a split point}
\fixme{``Why is the choice of the split point not part of this paper?''}

\hlc{GUIDE and MOB, which are both model-based approaches, employ the model estimated 
in the considered node for an exhaustive search of the best split point over all possible 
split points in the already selected split variable.
CTree has not been developed specifically for model-based partitioning, hence, a 
different approach to selecting a split point is suggested in the original paper. 
However, as an adaption of CTree employing models for the split variable selection 
is considered in this study, it is also combined with the same exhaustive search 
strategy as GUIDE and MOB.}
%\newpage

\subsection{Pruning}
\label{sec:pruning}

@@

\fixme{``Do the building blocks cover all possible options?''}

Note that there are further differences in the testing strategies between the three algorithms, e.g.,
using conditional vs.\ unconditional approximations of the null distributions. However, this difference
has relatively little influence compared to the other building blocks considered in detail here.
Moreover, both similarities and relative differences between these approaches have been previously
discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.
\hlc{Therefore, in this study the focus is put on the selected building blocks as they are assumed 
(and have shown) to have the highest influence on the difference in performance between the testing 
strategies.}

@@

%\newpage
\subsection[``Stump'' scenario: Residuals vs. full model scores]{``Stump'' scenario: Residuals vs.\ full model scores}
\label{sec:scores}

@@

However, in the scenario where only $\beta_1$ (but not the intercept $\beta_0$) is affected by the split,
the residual-based GUIDE approach has no power at all even for a true split at the median (top middle panel).
It is easily possible, though, to substantially mitigate this problem by using scores
(sensitive to changes in all parameters) rather than residuals only (sensitive to changes
in the conditional mean). The remaining difference between MOB/CTree and GUIDE+scores
is due to dichotomizing the scores at zero and due to categorizing the split variables
which are investigated in more detail in the next section.

\fixme{Type I error}

Figure~\ref{fig:scores_stump} \hlc{also shows that for $\delta = 0$, i.e., no difference
in the parameters and consequently a random split variable selection, the probability of 
selecting $Z_1$ as split variable is (close to)~$0$ (with the employed Bonferroni-adjustment of 
$p$-values keeping this value particularly low).
Even though it has to be noted that the number of replications in this simulation study is of
course limited, the results for this setting indicate that all testing strategies conform well with 
the predefined level of significance $\alpha = 0.05$ as the probability of splitting in any 
of the available covariates is between 4 and 5~percent for $\delta = 0$ 
(the exact values can be found in} Table~\ref{tab:typeIerror} in Appendix~\ref{app:typeIerror})
\hlc{with the tests for structural changes being slightly more conservative} 
\citep[see, e.g.,][for more details on this typical behaviour]{Hothorn+Zeileis:2008, Zeileis+Hothorn:2013, Hansen:1997}.

@@

Linear models as presented in this study have been chosen as they are highly relevant in practice,
allow for simple illustrations, and theoretical insights are available for the testing strategies.
However, the results can be easily extended to a wide variety of other models where the introduced building
blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
changes in the parameters do not lead to shifts in the residuals (but in other components of the
scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
of a linear model only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor which in our case is $(1,0)^{\top}$.
Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
is of type $(0, \delta)^\top$ and thus orthogonal to the mean regressor. Due to this residual-based
tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
interactions, see \citealp{Doove+Dusseldorp+VanDeun:2014}) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.
Testing strategies being capable of detecting all kinds of changes including those orthogonal to the 
mean are also employed in an extension of (generalised) linear model trees introduced by 
\cite{Seibold+Hothorn+Zeileis:2019}
which allows for selected model parameters to be kept fixed at a globally estimated value while only 
changes in the remaining parameters are considered for splitting.

@@

\section*{Acknowledgments}
Torsten Hothorn received funding from the Swiss National Science
Foundation, grant number 200021\_184603.
Lisa Schlosser received a PhD scholarship granted from the University of Innsbruck.

@@

For the evaluation of all 12 possible combinations of the presented building blocks
the original testing strategies CTree, MOB, and GUIDE have been applied together
with modified versions of them. The employed versions and the corresponding combination
of building blocks are listed in Table~\ref{tab:combinations_appendix}.

@@

\section{Type I error}
\label{app:typeIerror}
<<typeIerror, echo=FALSE, results=hide>>=
# Table: Type I error
subdata <- subset(sim_stump, test %in% c("ctree","mfluc",
                                         "guide_sum_12", "guide_sum_1_cor"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree","mfluc",
                                  "guide_sum_12", "guide_sum_1_cor"),
                       labels = c("CTree","MOB",
                                  "GUIDE+scores", "GUIDE"))

sd <- subset(subdata, changetype == "abrupt" & delta == 0 & xi %in% c("0","0.8"))
sd$xi <- factor(sd$xi)
# sd[, c("test","xi", "variation", "prop_split")]

xtbl <- xtabs(prop_split ~ (xi + test) + variation, data = sd)

# sd$prop_split[sd$test == "CTree" & sd$variation == "all" & sd$xi == "0"]
@

\begin{table}[h]
\centering
\begin{tabular}{ l l c c c }
\hline\noalign{\smallskip} 
Split point  & Strategy     & Varying $\beta_0$  & Varying $\beta_1$ & Varying $\beta_0$ and $\beta_1$ \\ 
\noalign{\smallskip}\hline\noalign{\smallskip} 
$\xi = 0$   & CTree         &  0.05 & 0.05 & 0.05 \\ 
($50$\%)    & MOB           &  0.04 & 0.04 & 0.04 \\
            & GUIDE+scores  &  0.05 & 0.05 & 0.05 \\
            & GUIDE         &  0.04 & 0.04 & 0.04 \\
\noalign{\smallskip}\hline\noalign{\smallskip} 
$\xi = 0.8$ & CTree         &  0.05 & 0.05 & 0.05 \\
($90$\%)    & MOB           &  0.04 & 0.04 & 0.04 \\
            & GUIDE+scores  &  0.05 & 0.05 & 0.05 \\
            & GUIDE         &  0.04 & 0.04 & 0.04 \\
\noalign{\smallskip}\hline
\end{tabular}
\caption{Probability of splitting in any of the available split variables,
i.e., probability of any of the available split variables showing a $p$-value
below the predefined level of significance $\alpha = 0.05$. 
The testing strategies CTree, MOB, GUIDE, and GUIDE+scores are considered in 
the ``stump'' scenario. Probabilities are estimated for effect size $\delta = 0$ 
(i.e., no difference in the parameters) with 100 replications of 250 observations.
The true split point $\xi$ in $Z_1$ is either the median~$0$~(top row) or the 90\%~quantile~$0.8$~(bottom row)
for either varying intercept~$\beta_0$, varying slope~$\beta_1$, or both coefficients varying.
\label{tab:typeIerror}}
\end{table}
